Sample Clustering

### 2017-02-02 Clustering.Rmd
Predict NP vs P using number of clusters

@@

patient.info$Patient = gsub("/", "_", patient.info$Patient)
head(patient.info)

@@

## How do the samples separate?

@@

```{r clust, echo=T, message=T, warning=F, eval=F}

for (pt in subset(sum.patient.data, Status == 'P')$Patient) {
  if (pt == 'AH0254') next
  print(pt)
  print(ncol(patient.data[[pt]]$seg.vals[,-(1:5)]))

  x1 = patient.data[[pt]]$seg.vals[,-(1:5)]

  model = Mclust(t(x1))
  print(model$loglik)

  tryCatch({
    mod1dr = MclustDR(model, lambda = 0)
    #summary(mod1dr)
    table(mod1dr$mixcomp)

    plot(mod1dr, what = "evalues")
    title(main=pt)
    plot(mod1dr, what = "density")
  })
}

```
```{r apclust-func, echo=F, message=F, warning=F}
suppressPackageStartupMessages( library(apcluster) )

@@

```{r em, echo=F, message=F, warning=F, eval=F} 
## EM

@@

## Perhaps AP Clustering

## Affinity Propagagion clustering described by Frey & Dueck, 2007

@@

```{r apclust, echo=T, message=T, warning=F, fig.height=8, fig.align='left'}
q=0.2

@@

  last = ncol(x1)
  if (patient.data[[pt]]$info$Status == 'P')
    last = sort(grep('HGD', patient.data[[pt]]$info$Pathology), decreasing=T)[1]
  if (ncol(x1) <= 15) {
#    plot(ac, x1[,c(c(1:ncol(x1))[-last], last)], main=pt)
  } else {
#    plot(ac, x1[,c(1, sort(sample(c(1:(ncol(x1)-1))[c(-1,-last)], 8)) ,last) ], main=pt)
  }

@@

@@

## (xval) Re run clusters on the Progressors without the final sample
There is still a difference between the progressors and non, though the p-value does increase. Some of the progressors do start looking more like NPs in their similarity measure but not all of them.

@@

for (pt in sum.patient.data$Patient) {
  x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename)
  if (subset(sum.patient.data, Patient == pt)$Status == 'P' & length(which(grepl('HGD', patient.data[[pt]]$info$Pathology))) > 0) {
    final = grep('HGD', patient.data[[pt]]$info$Pathology)
    x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename[-final])
  }
  ac = apcluster(negDistMat(x1, r=2), q=q)

@@

pander(subset(simX, Status == 'P' & ss >= mean(ss)), caption="These patients are more similar to Non-Progressors without their HGD sample(s)")
```

## Can we identify regions in the progressors?

Not really just from the SVDs.  This isn't any different at the moment from simply calling the variable regions directly.
```{r loc_pca, echo=F, warning=F, message=T, fig.height=12, fig.width=20}
count.occurrences<-function(grs) {
  ov = findOverlaps(grs, type="within", ignoreSelf=TRUE,select="arbitrary")
  counts = countOverlaps(grs, type='within', maxgap=0)

@@

>>>>

getPrecision<-function(actual,predicted) {
  retrieved <- sum(predicted==1)
  precision <- sum(predicted==1 & actual==1) / retrieved
  precision
}
getRecall<-function(actual,predicted) {
  retrieved <- sum(predicted==1)
  recall <- sum(predicted==1 & actual==1) / sum(actual==1)
  recall
}

@@

patient.info$Patient = gsub("/", "_", patient.info$Patient)
#head(patient.info)

@@

## How do the sample timepoints separate?

It appears that in the progressors that generally the HGD or IMC pathology may suddenly appear and be very different (PR1-HIN-044) from other samples, or those where it may have been a progressive sampling that can be fairly related to previous samples (AHM0952).  In most cases the HGD/IMC samples drive the variance, but it is often there in the LGD samples as well.  In contrast the non-progressor patients, even where they have LGD, don't generally show a sudden jump in the sample (AD0591).

This would suggest we might have a more difficult time separating out the progressors where the early samples were all highly related and a sudden shift occurred to generate the HGD sample. This could be either sampling bias or a sudden model of progression, but I can't think of a method right now that might separate the two.

@@

## Mixture Models

One of the questions we'd like to answer is how to separate the progressors from the non-progressors before we see the pathological progression (HGD or even LGD).  If there is increasing instability in the genome how do we characterize it?

I tried were several mixture models (e.g. EM) but wasn't able to find a pattern that could reliably separate the few mixtures I did see

```{r clust, echo=T, message=T, warning=F, eval=T, fig.height=6, fig.width=6}

llk = vector(mode='numeric', length=nrow(sum.patient.data))
names(llk) = sum.patient.data$Patient
for (pt in sum.patient.data$Patient) {
  print(pt)
  print(ncol(patient.data[[pt]]$seg.vals[,-(1:5)]))

  x1 = as.matrix(patient.data[[pt]]$seg.vals[,-(1:5)])
  x1 = x1[,patient.data[[pt]]$info$Samplename]

  fit <- Mclust(x1, G=3)
  plot(fit, what='classification')
  title(main=paste(pt, ' (',patient.data[[pt]]$info$Status[1], ')', sep=''))

  llk[pt] = fit$loglik

#   clusterPlots <- data.frame(x1 = clusterData$x1, x2 = clusterData$x2, cluster = factor(model$classification, 
#     levels = c(1, 2), labels = c("z = 0", "z = 1")), uncertainity = model$uncertainty, 
#     logUncertainity = log(model$uncertainty))
# clusterPlots.gg <- ggplot(clusterPlots)
# clusterPlots.gg + geom_point(aes(x = x1, y = x2, color = cluster))

  #tryCatch({
  # model = densityMclust(x1, G=3)
  # plotDensityMclustd(model, data=x1,  col='blue', lwd=2, hist.col = "lightblue",  breaks=100)
  #   mod1dr = MclustDR(model, lambda = 0.5)
  #   #summary(mod1dr)
  #   table(mod1dr$mixcomp)
  # 
  #   #plot(mod1dr, what = "evalues")
  #   #title(main=pt)
  #   plot(mod1dr, what = "density")
  #   title(main=paste(pt, ' (',patient.data[[pt]]$info$Status[1], ')', sep=''))
  # })
}

# Log liklihood doesn't separate them
t.test(llk[subset(sum.patient.data, Status == 'NP')$Patient], llk[subset(sum.patient.data, Status == 'P')$Patient])

```{r means, echo=F, message=F, warning=F, eval=F} 
## Kmeans
#   x1 = as.matrix(patient.data[[pt]]$seg.vals[,-(1:5)])
# 
   fit = kmeans(x1, 3)
#   plot(x1[,c(1,10)], col=km$cluster)
# title(main=paste(pt, ' (',patient.data[[pt]]$info$Status[1], ')', sep=''))

library(cluster) 
clusplot(x1, fit$cluster, color=TRUE, shade=TRUE, labels=2, lines=0, main=pt)

```{r apclust-func, echo=F, message=F, warning=F}
suppressPackageStartupMessages( library(apcluster) )

@@

@@

## AP Clustering

So tried AP as that doesn't require the pre-definition for number of clusters.  It appears to give good separation, though there's a suggestion of two different types of progressors.

### Affinity Propagagion clustering described by Frey & Dueck, 2007

@@

```{r apclust, echo=T, message=T, warning=F, fig.height=10, fig.width=10, fig.align='left'}
q=0.2

@@

  last = ncol(x1)
  if (patient.data[[pt]]$info$Status[1] == 'P' & pt != 'AD0252')
    last = sort(grep('HGD|IMC', patient.data[[pt]]$info$Pathology), decreasing=T)[1]
  if (ncol(x1) <= 15) {
    plot(ac, x1[,c(c(1:ncol(x1))[-last], last)], main=pt)
  } else {
    plot(ac, x1[,c(1, sort(sample(c(1:(ncol(x1)-1))[c(-1,-last)], 8)) ,last) ], main=pt)
  }

@@

fit = glm(Status~clusters, sim, family=binomial(link='logit'))
pander(fit, justify='left')

sim = sim[, c('Status', 'clusters', 'ns','ss')]
actualMatrix = matrix(0,1000,2,dimnames=list(c(), c('prec','recall')))
for (i in 1:nrow(actualMatrix)) {
  trainingRows = sample(1:nrow(sim), nrow(sim)-5)
  train = sim[trainingRows,]
  test = sim[-trainingRows,]

  fit = glm( (as.integer(Status)-1)~clusters, train, family=binomial(link='logit'))
  pred = predict(fit, test, se.fit=F) # probabilities -- of which outcome though?

  actual = test$Status == 'P'
  predicted = pred>0

  actualMatrix[i,'prec'] = getPrecision( actual, predicted )
  actualMatrix[i, 'recall'] = getRecall( actual, predicted )
}

ggplot(as.data.frame(actualMatrix), aes(x=recall, y=prec)) + geom_point() + 
  geom_hline(yintercept = 0.5, col='red', alpha=0.5) +
  labs(title='Precision/Recall for Progressor/NP prediction')

### Are the clusters random?

```{r example_pt}
pt = 'PR1_HIN_042'
patient.data[[pt]]$info = arrange(patient.data[[pt]]$info, Endoscopy.Year, desc(Pathology))

cols = c('chr','cluster','P.x', 'x')
hgmx = matrix(nrow=0,ncol=length(cols),dimnames=list(c(), cols))
for (chr in c(1:22)) {
  for (i in 1:length(ac@clusters)) {
    cl = x1[ac@clusters[[i]],]

    x = length(grep(paste('^',chr,'(p|q)',sep=''),  rownames(cl), value=T)) # n in cluster
    m = length(grep(paste('^',chr,'(p|q)',sep=''),  rownames(x1), value=T)) # total n
    n = nrow(x1)-m
    k = length(ac@clusters[[i]])

    hgmx = rbind(hgmx, cbind(chr, i, signif(dhyper(x,m,n,k),3), x))
  }
}

hgmx = as.data.frame(hgmx)

ggplot(subset(hgmx, x > 0), aes(x=chr,y=P.x,color=factor(cluster))) + ylim(0,1)+ 
  geom_point() + geom_line(alpha=0.1)

ggplot(subset(hgmx, x > 0 & cluster == 9), aes(x=factor(chr),y=P.x,fill=factor(cluster))) + ylim(0,0.5 )+ 
  geom_bar(colour='grey', stat='identity', position=position_dodge())

@@

## (xval) Re run clusters on the Progressors without the HGD sample
There is still a difference between the progressors and non, though the p-value does increase. Some of the progressors do start looking more like NPs in their similarity measure but not all of them.

@@

for (pt in sum.patient.data$Patient) {
  x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename)
  if (subset(sum.patient.data, Patient == pt)$Status == 'P' & length(which(grepl('HGD|IMC', patient.data[[pt]]$info$Pathology))) > 0) {
    final = grep('HGD|IMC', patient.data[[pt]]$info$Pathology)
    x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename[-final])
  }
  ac = apcluster(negDistMat(x1, r=2), q=q)

@@

pander(subset(simX, Status == 'P' & ss >= mean(ss)), caption="These patients are more similar to Non-Progressors without their HGD sample(s)")

simX = simX[, c('Status', 'clusters', 'ns','ss')]
noHGDPR = as.data.frame(matrix(0,1000,2,dimnames=list(c(), c('prec','recall'))))
for (i in 1:nrow(actualMatrix)) {
  trainingRows = sample(1:nrow(simX), nrow(simX)-5)
  train = simX[trainingRows,]
  test = simX[-trainingRows,]

  fit = glm( Status~clusters, train, family=binomial(link='logit'))
  pred = predict(fit, test, se.fit=F) # probabilities -- of which outcome though?

  actual = test$Status == 'P'
  predicted = pred>0

  noHGDPR[i,'prec'] = getPrecision( actual, predicted )
  noHGDPR[i, 'recall'] = getRecall( actual, predicted )
}

ggplot(as.data.frame(noHGDPR), aes(x=recall, y=prec)) + geom_point() + 
  geom_hline(yintercept = 0.5, col='red', alpha=0.5) +
  labs(title='Precision/Recall for Progressor/NP prediction NO HGD')

Currently, without the HGD/IMC samples we can predict progressors `r nrow(subset(noHGDPR, prec > 0.5))/nrow(noHGDPR)*100`% of the time.
In comparison, including HGD/IMC samples our predictions improve to `r length(which(actualMatrix[,'prec'] > 0.5))/nrow(actualMatrix)*100`%.

## And why q=0.2
```{r whyq, echo=F, message=F, warning=F}
sum.patient.data = arrange(sum.patient.data, Status)

pvals = as.data.frame(matrix(ncol=2,nrow=0,dimnames=list(c(), c('clusters','ss'))))
for (qv in seq(0.1,0.9,0.1)) {
  print(qv)

  ## Using the segmentation values
  simPC = sum.patient.data[,c(2:4)]
  rownames(simPC) = sum.patient.data[,1]
  for (pt in sum.patient.data$Patient) {
    patient.data[[pt]]$info = arrange(patient.data[[pt]]$info, Endoscopy.Year, desc(Pathology))

    ac = apcluster(negDistMat(x1, r=2), q=qv, convits=75)

    simPC[pt, 'ns'] = net.similarity(ac)
    simPC[pt, 'ss'] = sum.similarity(ac)
    simPC[pt, 'clusters'] = length(clusters(ac))
  }

  # May be a difference
  ttps = t.test( subset(simPC, Status == 'NP')$ss, subset(simPC, Status == 'P')$ss )$p.value
  ttpn = t.test( subset(simPC, Status == 'NP')$ns, subset(simPC, Status == 'P')$ns )$p.value
  ttpc = t.test( subset(simPC, Status == 'NP')$clusters, subset(simPC, Status == 'P')$clusters )$p.value

  pvals = rbind(pvals,cbind(ttpc,ttps))
}
colnames(pvals) = c('clusters','sum.similarity')
pvals$qv=seq(0.1,0.9,0.1)

ggplot(melt(pvals, id.vars='qv'),aes(x=qv,y=value, color=variable)) + geom_point() +
  labs(main="qvalus for AP", y='pvalue NP vs P', x='q param')

```{r loc_pca, echo=F, warning=F, message=T, fig.height=12, fig.width=20, eval=F}
## Can we identify regions in the progressors?

#Not really just from the SVDs.  This isn't any different at the moment from simply calling the variable regions directly.

count.occurrences<-function(grs) {
  ov = findOverlaps(grs, type="within", ignoreSelf=TRUE,select="arbitrary")
  counts = countOverlaps(grs, type='within', maxgap=0)

@@

```{r precrec, echo=F, warning=F, message=F, fig.height=12, fig.width=20}
